---
title: "Mine Waste in the U.S."
subtitle: "A Geological Analysis using USGS NMWI Data"
author: "Indigo Sterud"
institute: "University of Illinois at Chicago"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["default", "xaringan-themer.css"]
    nature:
      ratio: "16:9"
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
---
```{r load-packages, include = FALSE}
# Add any additional packages you need to this chunk
library(ggplot2)
library(xaringan)
library(tidyverse)
library(tidymodels)
library(palmerpenguins)
library(knitr)
library(xaringanthemer)
library(readr)
library(kableExtra)
library(purrr)
library(here)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_solarized_light(text_color = "gray31",
  header_color = "#dc322f",
  background_color = "#fdf6e3",
  link_color = "#b58900",
  text_bold_color = "#d33682",
  text_slide_number_color = "#93a1a1")
```

```{r setup, include=FALSE}
# For better figure resolution
knitr::opts_chunk$set(fig.retina = 3, dpi = 300, fig.width = 6, fig.asp = 0.618, out.width = "80%")
```

```{r load-data, include=FALSE}
Waste_Geology <- read_csv(here::here("project-starter", "data", "Waste_Geology.csv"))
Waste_Resources <- read_csv(here::here("project-starter", "data", "Waste_Resources.csv"))

mining_df <- left_join(Waste_Resources, Waste_Geology, by = "Site_ID") %>%
  rename(
    Site_Name = Site_Name.x,
    Site_Name_geo = Site_Name.y)

mining_clean <- mining_df %>%
  mutate(
    state_abbr = substr(Site_ID, 1, 2),
    state_name = state.name[match(state_abbr, state.abb)]
  ) %>%
  filter(!is.na(Calc_Vol), Calc_Vol > 0) %>%
  mutate(log_vol = log10(Calc_Vol)) %>%
  drop_na(Calc_Vol, Ore_Min)
```


## Research Questions

- Are there differences in waste production across states?

--

- Do certain mineral commodities produce more waste?

---

## Mining Data
- Columns represent different mine waste features for each mine location.
- Commodity = Minerals mine factories want
- Gangue_Min = non-commodity minerals

```{r glimpse-table, echo = FALSE}
mining_clean %>%
  select(Site_Name, state_abbr, Calc_Vol, Commodity, Gangue_Min) %>%
  head(10) %>%
  kable(format = "html") %>%
  kable_styling(
    full_width = FALSE,
    font_size = 10
  )
```

---
## Waste Volume Basics
```{r extra-stuff, echo = FALSE}
ggplot(mining_clean %>% filter(Calc_Vol > 0), aes(x = Calc_Vol)) + 
  geom_histogram(bins = 20) + 
  scale_x_log10() + 
  theme_bw() +
  labs(
    title = "Distribution of Waste Volumes",
    x = "log10(Waste Volume (m^3))",
    y = "Count"
  )
```

---

class: center, middle

## Differences in Calculating Mining Waste per State

---
class: center, middle
**Arizona and Utah have large outliers...**

This plot does not calculate for log(WasteVolume), showing the extremely large outliers in places like AZ and UT, creating a strong right skew to all of the data. For this reason, I used log scaling on all of my statistical analyses.

```{r state-waste-regular, echo = FALSE}
ggplot(mining_clean, aes(x = reorder(state_abbr, Calc_Vol), y = Calc_Vol)) + 
  geom_boxplot(fill = "steelblue") + 
  theme_bw(base_size = 12) + 
  labs( title = "Mine Waste Volume Distribution by State", x = "State", y = "Waste Volume (m^3)" )
```

---
##Using a log scale? Yes. Much Better.

```{r state-waste-log-plot, echo = FALSE, warning = FALSE}
ggplot(mining_clean, aes(x = reorder(state_abbr, Calc_Vol),y = Calc_Vol)) +
  geom_boxplot(fill = "pink") +
  scale_y_log10() +
  theme_bw(base_size = 10) +
  labs( title = "Mine Waste Volumes by State (Log Scale)", x = "State", y = "Waste Volume(m^3)")
```

---
## Other notes on mining data

1.) It is a giant compilaton of PUBLIC mining data across the US, compiled by USGS.


2.) Does not represent the whole of mines across each state / resource, especially private companies.


3.) This already gives the data a lot of room for error.


---

**Bootstrapping** helps since some states have more mining sites than others.

```{r state-waste-bootstrap, warning = FALSE, echo = FALSE}
set.seed(100)
bootstrap_state <- mining_clean %>%
  group_by(state_abbr) %>%
  summarize(
    boot_median = list(
      replicate(500,
        median(sample(Calc_Vol, replace = TRUE), na.rm = TRUE))
    )
  )

bootstrap_ci <- bootstrap_state %>%
  mutate(
    median_est = map_dbl(boot_median, median),
    lower = map_dbl(boot_median, ~quantile(.x, 0.025)),
    upper = map_dbl(boot_median, ~quantile(.x, 0.975))
  )

ggplot(bootstrap_ci,aes(x = reorder(state_abbr, median_est), y = median_est)) +
  geom_col(fill = "steelblue") +
  scale_y_log10() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.35) +
  theme_bw(base_size = 12) +
  labs(
    title = "Bootstrapped Median Mine Waste Volume by State",
    x = "State",
    y = "Median Volume"
  )
```
---
## ANOVA test Waste Volumes by State
Using a logarithmic calculation, the Pr(>F) of <2e-16 reveals that there is at least one state of 10 that differ significantly from the others. Also, the large F-value (107.1) indicates that differences between states are much greater than variability within individual states.

```{r anova-state, echo=FALSE, warning=FALSE}
anova_model_1 <- aov(log_vol ~ state_abbr, data = mining_clean)

summary(anova_model_1)
```
---
class: inverse, middle, center

## Commodity Minerals and Waste

What is the relationship between the type of mineral and the amount of waste produced?

---
```{r commodity-data, echo=FALSE}

Commodity_Long <- mining_clean %>%
  separate_rows(Commodity, sep = ",|;") %>%
  mutate(
    Commodity = trimws(Commodity),
    Commodity = gsub("\\s*\\(.*\\)", "", Commodity)
  )

Top20 <- Commodity_Long %>%
  group_by(Commodity) %>%
  summarize(median_vol = median(Calc_Vol, na.rm = TRUE)) %>%
  slice_max(median_vol, n = 20)

```

```{r commodity-plot, echo=FALSE}

ggplot(Top20, aes(x = reorder(Commodity, median_vol), y = median_vol)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Median Waste Volume for Commodity Minerals Identified",
    x = "Commodity Mineral",
    y = "Median Volume (m^3)"
  )

```

- Tellurium is often harvested from large Mine Waste Locations

- Gold is often associated with mines that are smaller scale and produce less waste.

---
# Commodity Scatterplot

**Silver, Molybdenum, Gold and Copper** are the most prevalent.

```{r commodity-scatter, echo=FALSE, warning=FALSE}

Commodity_Filtered <- Commodity_Long %>%
  filter(Commodity %in% Top20$Commodity, Calc_Vol > 0)

ggplot(Commodity_Filtered, aes(x = Commodity, y = Calc_Vol, color = Commodity)) +
  geom_jitter(alpha = 0.6, width = 0.2) +
  coord_flip() +
  scale_y_log10() +
  theme_bw() +
  theme(text = element_text(family = "serif"),
        legend.position = "none") +
  labs(
    title = "Mine Waste Volume by Commodity Mineral (Individual Sites)",
    x = "Commodity Mineral",
    y = "Waste Volume (m^3)"
  )

```
---

1.) The most amount of mine locations contained copper, the least amount of mine locations contained silver.

2.) All had roughly the same shape, probably meaning they were found many of the same sites.

3.) All had roughly the same median amount of waste volume at the sites they were harvested (1E07)

```{r commodity-hist, echo=FALSE, warning=FALSE}
ggplot(
  Commodity_Long %>%
    mutate(Commodity = trimws(Commodity)) %>%
    filter(Commodity %in% c("gold", "copper", "molybdenum", "silver"), Calc_Vol > 0),
  aes(x = log10(Calc_Vol), fill = Commodity)) +
  geom_histogram(alpha = 0.35, bins = 15) +
  theme_bw(base_size = 10) +
  scale_fill_manual(values = c(
    "gold" = "#E69F00",
    "copper" = "#D55E00",
    "silver" = "grey",
    "molybdenum" = "steelblue")) +
  labs(
    title = "Distribution of Mine Waste Volume by Commodity",
    x = "log10(Waste Volume (m^3))",
    y = "Count",
    fill = "Commodity")

```
---
## ANOVA test for Commodity Minerals

Comparing 49 Commodity minerals, Pr(<2e-16) shows that the difference in means between individual groups is extremely statistically significant, meaning different mineral commodities have different mean waste volume amounts, although they may be found at the same sites.
F-value (21.93) shows that the variation between commodity mineral means is significantly larger than the variation within the groups themselves.

```{r anova-commodity, echo=TRUE, warning=FALSE}
anova_model <- aov(
  log10(Calc_Vol) ~ Commodity,
  data = Commodity_Long
)

summary(anova_model)
```

---
## T-test: Gold and Copper
No statistically significant difference in average log-transformed mine waste volume sites between copper and gold (p = 0.2922).

```{r t-test gc, echo = FALSE}
t.test(
  log_vol ~ Commodity,
  data = mining_clean %>%
    filter(Commodity %in% c("copper", "gold"))
)
```
---
## T-test: Copper and Molybenum
p = 1.531e-05 shows a highly significant difference between copper and molybdenum waste sites.

- Molybdenum sites are found in mines with much higher waste volumes than copper.
```{r gold and molybdenum, echo = FALSE}
t.test(
  log_vol ~ Commodity,
  data = mining_clean %>%
    filter(Commodity %in% c("copper", "molybdenum"), Calc_Vol > 0)
)
```
---

# Thanks!
